A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-18 sections/Overview.tex
Definition of declarative terms

\begin{figure}[t]
\[
\begin{array}{l@{\qquad}lcl}
\text{Type variables}\qquad&a, b\\[3mm]
\text{Types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \forall a. A \mid A\to B\\
\text{Monotypes}\qquad&\tau,\sigma &::=&\quad 1 \mid a \mid \tau\to \sigma\\
\text{Contexts}\qquad&\Psi &::=&\quad \nil \mid \Psi, a\\
%\text{Judgments}\qquad&\exps &::=&\quad \nil \mid A \le B : \exps
\end{array}
\]
\caption{Syntax of Declarative System}\label{fig:decl:syntax}
\end{figure}

The syntax of this declarative system is shown in Figure~\ref{fig:decl:syntax}.
A declarative type $A$ is either the unit type $1$, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.
Nested universal quantifiers are allowed in types,
but monotypes $\tau$ do not contain any universal quantifier.
Contexts $\Psi$ are just sequences of type variable declarations.

@@

Figures~\ref{fig:decl:sub} and \ref{fig:decl:typing},
give the subtyping and typing relations for the declarative DK system.
The DK subtyping relation was adopted from \citet{odersky1996putting}.

\paragraph{Declarative Subtyping}
Figure~\ref{fig:decl:sub} shows DK's declarative subtyping judgment $\Psi \vdash A \le B$,
which was adopted from \citet{odersky1996putting}. It compares the
degree of polymorphism between between $A$ and $B$ in DK's implicit polymorphic type system. 
Essentially, if $A$ can always be instantiated to match any instantiation of $B$,
then A is ``at least as polymorphic as'' $B$. We also 
say that $A$ is ``more polymorphic than'' $B$ and write $A \le B$.

Subtyping rules $\mathtt{{\le}Var}$, $\mathtt{{\le}Unit}$ and $\mathtt{{\le}{\to}}$
handle simple cases that do not involve universal quantifiers.
The subtyping rule for function types $\mathtt{{\le}{\to}}$ is standard,
being covariant on the return type and contravariant on the argument type.
Rule $\mathtt{{\le}\forall R}$ states that if $A$ is a subtype of $B$
in the context $\Psi, a$, where $a$ is fresh in $A$, then $A\le\all B$.
Intuitively, if $A$ is more general than $\all B$ (where the universal quantifier
already indicates that $\all B$ is a general type),
then $A$ must instantiate to $[\tau/a]B$ for every $\tau$.

The most interesting rule is $\mathtt{{\le}\forall L}$.
If some instantiation of $\all A$, $[\tau/a]A$, is a subtype of $B$,
then $\all A \le B$.
The monotype $\tau$ we used to instantiate $a$ is \emph{guessed} in this
declarative rule, but the algorithmic system does not guess and defers the
instantiation until it can determine the monotype deterministically.
The fact that $\tau$ is a monotype rules out the possibility of
polymorphic (or impredicative) instantiation.\tom{say why this restriction is imposed}
Peyton Jones et al.~\cite{} also impose the restriction of
predicative instantiation in their type system.
Both systems are adopt by several practical programming languages.

>>>>

\begin{figure}[t]
\[
\begin{array}{l@{\qquad}lcl}
\text{Type variables}\qquad&a, b
\\[3mm]
\text{Types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \forall a. A \mid A\to B\\
\text{Monotypes}\qquad&\tau,\sigma &::=&\quad 1 \mid a \mid \tau\to \sigma
\\[3mm]
\text{Expressions}\qquad&e &::=&\quad x \mid () \mid \lam e \mid e_1~e_2 \mid (e:A)
\\[3mm]
\text{Contexts}\qquad&\Psi &::=&\quad \nil \mid \Psi, a \mid \Psi, x:A
\end{array}
\]
\caption{Syntax of Declarative System}\label{fig:decl:syntax}
\end{figure}

The syntax of this declarative system is shown in Figure~\ref{fig:decl:syntax}.
A declarative type could be a unit type 1, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.
Nested universal quantifiers are allowed for types,
but monotypes do not have any universal quantifier.
Terms include a unit term $()$, variables $x$, lambda-functions $\lam e$,
applications $e_1~e_2$ and annotations $(e:A)$.
Context $\Psi$ is a collection of type variable declarations and
term variables with its type declared $x:A$.

@@

In Figures~\ref{fig:decl:sub} and \ref{fig:decl:typing},
we give the subtyping and typing relations for the declarative system,
which is exactly the same as DK's.
The subtyping relations originate from Odersky and La\"ufer~\cite{odersky1996putting},
and is adopted by DK as a part of their typing system.

\paragraph{Declarative Subtyping}
For implicit polymorphic type systems,
subtyping relation compares the degree of polymorphism of types.
In short, if a polymorphic type $A$ always succeed to instantiate to any instantiation of $B$,
then A is ``at least as polymorphic as'' $B$,
or we just say that $A$ is ``more polymorphic than'' $B$, written as $A \le B$.

Subtyping rules $\mathtt{{\le}Var}$, $\mathtt{{\le}Unit}$ and $\mathtt{{\le}{\to}}$
handle simple cases where no universal quantifiers get involved.
The subtyping rule for function types $\mathtt{{\le}{\to}}$ is standard,
being covariant on the return type and contravariant on the argument type.
Rule $\mathtt{{\le}\forall L}$ states that if $A$ is a subtype of $B$
under the context ($\Psi, a$), where $a$ is fresh in $A$, then $A\le\all B$.
Intuitively, if $A$ is more general than $\all B$ (where the universal quantifier
already indicates that $\all B$ is a general type),
then $A$ must instantiate to $[\tau/a]B$ for every $\tau$.

The most interesting rule is $\mathtt{{\le}\forall R}$.
If any instantiation of $\all A$, $[\tau/a]A$, is a subtype of $B$,
then $\all A \le B$.
The monotype $\tau$ we used to instantiation is \emph{guessed} for the declarative rule,
but the algorithmic system should be able to compute the guess properly.
The fact that $\tau$ is a monotype rules out the possibility of
polymorphic (or impredicative) instantiation.
Peyton Jones et al.~\cite{} also has such restriction of
predicative instantiation in their type system.
Both systems are adopt by several practical programming languages.
